Quantum Semiprime Factorization: Leveraging Grover's Algorithm for Efficient Prime Decomposition

### 2025-03-02 LaTeX/paper.tex
Talked about shor's and added figures to paper

% \author {Jacob Collins$^{1}$, Jaime Raigoza$^{1}$, Sam Siewert$^{1}$ \\
\author {Jacob Collins$^{1}$, Jaime Raigoza$^{1}$, Sam Siewert$^{1}$ \\
\ \\ % leave an empty line between authors and affiliation
 $^1$ California State University, Chico, United States of America (note- full address at end of paper)}

@@

Grover's algorithm has an incredibly wide range of potential use-cases,
and an intention for this paper is to provide a generalized reference point
for how Grover's algorithm may be used to invert any function.

@@

  Given a function $f(x)=y$, where $x$ is unknown (index, prime factors, 
  sum components, etc.), and $y$ is known (array value, semiprime/product, 
  sum, etc.), Grover's algorithm effectively takes the role of $f'(y)$, 
  allowing for a potential speedup in finding whatever input(s) to $f(x)$
  will return $y$.

@@

  Description of Shor's algorithm.

@@

  S. Whitlock, et al. present a method of SP factoring with Grover's
  algorithm. The implementation in their paper is highly optimized,
  and modifies the target value in their circuit from some semiprime $N$,
  to $M$, a reduced number uniquely tied to $N$, which has prime factors
  $p$ and $q$.

  While this approach is admirable, it introduces a level of complexity 
  that may hinder a learner's understanding of the mechanics at play,
  so the implementation shown in section \ref{sec:Methodology} forgoes
  this abstraction from $N$ to $M$, and instead implements an oracle which
  searches directly for prime factors $a$ and $b$ of a semiprime $N$.

\section{Methodology} \label{sec:Methodology}

@@

  Mention unitary matrices.

  Describe how Grover's works and show figures of a generalized
  circuit for Grover's to invert functions

  Reference figure 1.

  Operations can be performed on a register in superposition, and these operations
  will affect each potential state individually. This property is utilized by
  Grover's Oracle to selectively target specific states which match the target
  value, and the diffuser is used to increase the probability of measuring the
  inputs which resulted in this marked target state.

  \begin{figure}[h]
  %\centering
  \includegraphics[width=7.9cm]{grover_inversion.png}
  \caption{General circuit diagram for Grover's Algorithm, used to find 
  the inputs $x, y$ for which $f(x,y)=z$, for some known $z$.}
  \label{FIGURA1}
  \end{figure}

  Include one figure showing the oracle, and one figure showing the diffuser.

@@

  Describe our original approach- bitwise addition, registers, operations, etc.

  Mention why an alternative approach was needed.

  Include a drawing of the circuit.

>>>>

% \author {Jacob Collins$^{1}$, Jaime Raigoza$^{1}$, Sam Siewert$^{1}$ \\
\author{Jacob Collins$^{1}$, Jaime Raigoza$^{1}$, Sam Siewert$^{1}$ \\
\ \\ % leave an empty line between authors and affiliation
 $^1$ California State University, Chico, United States (note- full address at end of paper)}

@@

Grover's algorithm has an incredibly wide range of potential use-cases,
and an intention for this paper is to provide a generalized reference
for how Grover's algorithm may be used to invert a given function.

@@

  Given a function $f(x)=y$, where $x$ is unknown (index, prime factors, 
  sum components, etc.), and $y$ is known (array value, semiprime/product, 
  sum, etc.), Grover's algorithm effectively takes the role of $f'(y)=x$, 
  allowing for a potential speedup in finding whatever input (s) to $f(x)$
  will return $y$, provided that it is much faster to compute $f(x)$ than
  whatever classical methods might be used to otherwise solve for $x$ given $y$.

@@

  Shor's algorithm is the most well-known approach to prime factorization
  in quantum computing. However, there are many drawbacks to this approach,
  the most notable of which being the need to take measurements mid-circuit
  and essentially re-run with new values afterwards, which introduces a 
  significant amount of overhead. Even recently improved versions of Shor's
  algorithm still require a minimum of $2n+1$ qubits, and tend to be 
  extremely sensitive to noise, which leads to high error\cite{quantum_factoring,shor}.

  Furthermore, when it comes to more concrete applications of Shor's algorithm,
  a full-scale implementation of this algorithm to factor an $n$-bit number
  may require up to $5n+1$ qubits for accurate results, and require 
  on the order of $n^3$ operations~\cite{quantum_fac_efficient}.

@@

  S. Whitlock, et al.\ present a method of SP factoring with Grover's
  algorithm. The implementation in their paper is highly optimized,
  and modifies the target value in their circuit from some semiprime $N$,
  to $M$, a reduced number uniquely tied to $N$, which has unique factors 
  $p$ and $q$ which can be used to calculate $a, b$, the prime factors of 
  $N$. This optimization from $N$ to $M$ allows the implementation to ignore
  the trivial prime factors $2$ and $3$, and requires fewer qubits than would
  otherwise be needed to search for prime factors of $N$~\cite{quantum_factoring}.

  While this approach is admirable, it introduces a level of complexity 
  that may hinder a learner's understanding of the mechanics at work,
  so the implementation shown in section\ \ref{sec:Methodology} forgoes
  this abstraction from $N$ to $M$, and instead implements an oracle which
  searches directly for prime factors $a$ and $b$ of a semiprime $N$.

\section{Methodology}\label{sec:Methodology}

\begin{figure*}[!ht]\label{fig:FIGURA1}
\centering
% \includegraphics[width=7.9cm]{grover_inversion.png}
\includegraphics[width=15.8cm]{grover_inversion.png}
\caption{General circuit diagram for Grover's Algorithm, used to find 
the inputs $x, y$ for which $f(x,y)=z$, for some known output $z$.}
\end{figure*}

@@

  Operations can be performed on a set of input registers that start in uniform 
  superposition, and these operations will affect each potential state individually. 
  The properties of superposition are utilized by Grover's Oracle to selectively 
  modify the states $|x,y\rangle$ in the input registers $R_x$ and $R_y$,
  marking them as valid by setting them to $-|x,y\rangle$, only when the output register
  $R_z$ matches some known search term $z$. An outline of a quantum circuit for
  Grover's algorithm is seen in Figure~\ref{fig:FIGURA1}.

  The oracle in Grover's algorithm performs a controlled operation which is meant
  to differentiate states $x,y$ that do result in the wanted target $z$ from 
  states $x,y$ that do not result in the target $z$.

  \begin{figure}[h]\label{fig:FIGURA2}
  \centering
  \includegraphics[width=6.0cm]{oracle_15.png}
  % \includegraphics[width=7.9cm]{oracle_15.png}
  \caption{Example circuit diagram for Grover's oracle marking states of the input
  registers $R_x$ and $R_y$ ($[q_0,q_2]$ and $[q_3,q_5]$, respectively) which
  result in $z=15$.}
  \end{figure}

  Figure~\ref{fig:FIGURA2} shows an example of the oracle searching for the state $z=15$,
  or rather, $R_z=|001111\rangle$. To implement a circuit which can selectively target
  $z=15$, Toffoli gates (otherwise known as NOT, or X gates) are applied on 
  $R_z[0]$ and $R_z[1]$, so that states where $R_z=|001111\rangle$ will temporarily be in
  state $R_z=|111111\rangle$, and the multi-controlled Toffoli gates\cite{multi_toffoli}
  will apply only for the desired states. These operations are then performed again,
  returning all but the target qubit (which has the Z-gate applied) to their original
  states.

  \begin{figure}[h]\label{fig:FIGURA3}
  \centering
  % \includegraphics[width=7.9cm]{diffuse.png}
  \includegraphics[width=5.0cm]{diffuse.png}
  \caption{Example circuit diagram for a diffuser which amplifies the probability that
  measurement will result in a state which has been marked by the oracle.
  Input registers $R_x$ and $R_y$ (q0-q2 and q3-q5, respectively) after diffusion are 
  more likely to be measured in a state such that $f(R_x,R_y)$ results in $R_z=|z\rangle$.}
  \end{figure}

  The diffuser in Grover's algorithm (Figure~\ref{fig:FIGURA3}) can then increase the 
  probability of observing $-|x,y\rangle$ (states which have been marked by the oracle) 
  through a process known as inversion about the mean.

@@

  As a proof of concept (prior to the release of~\cite{quantum_factoring}),
  a function $f(x,y)=x+y$ is defined such that $R_x=|x\rangle$, $R_y=|y\rangle$,
  and after performing $f(R_x,R_y)$, the output register $R_z$ shall be in the
  state $|x+y\rangle$.

  This approach was implemented in a similar manner to a classical full-adder,
  and worked on quantum simulators with $100\%$ accuracy~\cite{quantum_full_adder}.

  The semiclassical full-adder indeed works with Grover's algorithm, states were
  filtered to only measure inputs which resulted in a sum matching some given value.
  This is a trivial problem however, and only has potential usefulness due to the 
  relationship between multiplication and addition.

  Semiclassical arithmetic does have significant drawbacks, however. Despite
  being relatively simple to implement due to the similarity with principles
  that have a much broader range of documentation, the classical nature of this
  implementation meant that in order to move to multiplication from this point,
  the value of one register would need to be measured, and $R_x$ would then be
  added to $R_z$ a number of times dependent on the value of $y$.

  Measuring the state of either input register would collapse the superposition,
  and break the underlying principles on which Grover's algorithm depends. Thus,
  a different approach became necessary- one that could bridge the gap between 
  addition and multiplication without needing to measure either register,
  so that the operation can still work with inputs that are in superposition,
  allowing the use of Grover's algorithm.
